Proofs as Programs in Classical Logic\\Notes

### 2021-09-06 Notes/notes.tex
Introduced Agda

@@

For now we will consider two minimal examples. The first on doesn't use induction. There are two constant symbols $a, b$ and a binary proposition variable $p(-, -)$. The only axiom is $\forall x\:p(x,a)\vee p(x, b)$. We seek to prove $\forall x\exists y p(x, y)$. It is immediately clear that unless the axiom has computational content, our extracted program cannot yield a value. On the other hand we will see that we can still create a well-typed program. And if the axiom contains computational content it will indeed yield a result.

@@

Now for our computational translation we introduce a special exception terms $e_\alpha(t_1,\dots,t_n)$ and $\overline e_\alpha(t_1,\dots,t_n)$ for every atom $\alpha[t_1,\dots,t_n]$ with $n$ term-variables, which intuitively indicate that $\alpha[t_1,\dots,t_n]$ is not false. They are usually introduced in the manner of the last derivation of definition 3.4 in~\cite{de_Groote_1995}. If one were to assign types we would have $e_\alpha(t_1.\dots,t_n) : (\alpha(t_1,\dots,t_n)\to\bot)\to\bot$ and $\overline e_\alpha(t_1.\dots,t_n) : ((\alpha(t_1,\dots,t_n)\to\bot)\to\bot)\to\bot$. However as we have the (classical) identification, we can simplify $\overline e_\alpha(t_1.\dots,t_n) : \alpha(t_1,\dots,t_n)\to\bot$. Having this in mind both $e_\alpha(t_1.\dots,t_n) \overline e_\alpha(t_1.\dots,t_n) : \bot$ and $\overline e_\alpha(t_1.\dots,t_n) e_\alpha(t_1.\dots,t_n) : \bot$ make ``type-sense" in the exception sense of~\cite{de_Groote_1995}. We will use this in our program. Let us now translate the above proof. Note that the terms for axioms are given. For clarity we will use braces for function application. For convenience we will write $\alpha_i$ for the $i$-the generated term.
\setcounter{equation}{0}
\begin{example}\hfill
	\begin{align}
		& \alpha_1 & (p(x_1,a)\to\bot)\to(p(x_1, b)\to\bot)\to\bot\\
		& \overline e_p(x_2, y_2) & p(x_2, y_2)\to\bot\\
		& \lambda t. \mathbf{let}\:v : \neg p(x_1, a)&(p(x_1, b)\to\bot)\to\bot\\
		& \phantom{\lambda t.}\mathbf{in}\:\alpha_1(v, t) &\\
		& \phantom{\lambda t.}\mathbf{handle}\:v(w)\Rightarrow \overline e_p(x_1, a)(w)\indent [= \alpha_2[x_1/x_2, a/y_2](w)]\hspace*{-4cm}\nonumber&\\
		& \mathbf{let}\:v : \neg p(x_1, b)\:&\bot\\
		& \mathbf{in}\:\alpha_3(v) &\nonumber\\
		& \mathbf{handle}\:v(w)\Rightarrow \overline e_p(x_1, b)(w)\indent[= \alpha_2[x_1/x_2, b/y_2](w)]\hspace*{-4cm}\nonumber&\\
		& \lambda x_1.\alpha_4 &\forall x\bot
	\end{align}
\end{example}

Now to retrieve a value we just need to handle the final exception, i.e.
$$\lambda x_1.\alpha_4\:\mathbf{handle}\: \overline{e}_p(x_2, y_2)(w)\Rightarrow y_2$$
Of course this will not yield an actual value unless $\alpha$ yields.\\
We will now consider a minimal inductive example.\pagebreak

>>>>

\subsection{Type system}

Since we want to apply our method to first-order resolution proofs we will require a dependant type system. For this purpose we shall use the practical type system of agda as described in~\cite{agda}.

@@

For now we will consider two minimal examples. The first doesn't use induction. There are two constant symbols $a, b$ and a binary proposition variable $p(-, -)$. The only axiom is $\forall x\:p(x,a)\vee p(x, b)$. We seek to prove $\forall x\exists y p(x, y)$. It is immediately clear that unless the axiom has computational content, our extracted program cannot yield a value. On the other hand we will see that we can still create a well-typed program. And if the axiom contains computational content it will indeed yield a result.

@@

Now for our computational translation we introduce a special exception terms $e_\alpha(t_1,\dots,t_n)$ and $\overline e_\alpha(t_1,\dots,t_n)$ for every atom $\alpha[t_1,\dots,t_n]$ with $n$ term-variables, which intuitively indicate that $\alpha[t_1,\dots,t_n]$ is not false. They are usually introduced in the manner of the last derivation of definition 3.4 in~\cite{de_Groote_1995}. If one were to assign types we would have $e_\alpha(t_1.\dots,t_n) : (\alpha(t_1,\dots,t_n)\to\bot)\to\bot$ and $\overline e_\alpha(t_1.\dots,t_n) : ((\alpha(t_1,\dots,t_n)\to\bot)\to\bot)\to\bot$. However as we have the (classical) identification, we can simplify $\overline e_\alpha(t_1.\dots,t_n) : \alpha(t_1,\dots,t_n)\to\bot$. Having this in mind both $e_\alpha(t_1.\dots,t_n) \overline e_\alpha(t_1.\dots,t_n) : \bot$ and $\overline e_\alpha(t_1.\dots,t_n) e_\alpha(t_1.\dots,t_n) : \bot$ make ``type-sense" in the exception sense of~\cite{de_Groote_1995}. We will use this in our program. Let us now translate the above proof. Note that the terms for axioms are given. For clarity we will use braces for function application. For convenience we will write $\alpha_i$ for the $i$-the generated term.
\setcounter{equation}{0}
\begin{example}\hfill
	\begin{align}
		& \alpha_1 & (p(x_1,a)\to\bot)\to(p(x_1, b)\to\bot)\to\bot\\
		& \overline e_p(x_2, y_2) & p(x_2, y_2)\to\bot\\
		& \lambda t. \mathbf{let}\:v : \neg p(x_1, a)&(p(x_1, b)\to\bot)\to\bot\\
		& \phantom{\lambda t.}\mathbf{in}\:\alpha_1(v, t) &\nonumber\\
		& \phantom{\lambda t.}\mathbf{handle}\:v(w)\Rightarrow \overline e_p(x_1, a)(w)\indent [= \alpha_2[x_1/x_2, a/y_2](w)]\hspace*{-4cm}\nonumber&\\
		& \mathbf{let}\:v : \neg p(x_1, b)\:&\bot\\
		& \mathbf{in}\:\alpha_3(v) &\nonumber\\
		& \mathbf{handle}\:v(w)\Rightarrow \overline e_p(x_1, b)(w)\indent[= \alpha_2[x_1/x_2, b/y_2](w)]\hspace*{-4cm}\nonumber&\\
		& \lambda x_1.\alpha_4 &(x : X) \to \bot
	\end{align}
\end{example}

Now to retrieve a value we just need to handle the final exception, i.e.
$$\lambda x_1.\alpha_4\:\mathbf{handle}\: \overline{e}_p(x_2, y_2)(w)\Rightarrow y_2$$
Of course this will not yield an actual value unless $\alpha_1$ yields.\\
We will now consider a minimal inductive example.\pagebreak
